---
title: "Getting started with measr"
output: rmarkdown::html_vignette
bibliography: bib/references.bib
csl: bib/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with measr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of measr is to make it easy to estimate and evaluate diagnostic classification models (DCMs).
DCMs are primarily useful for assessment or survey data where responses are recorded dichotomously (e.g., right/wrong, yes/no) or polytomously (e.g., strongly agree, agree, disagree, strongly disagree).
When using DCMs, the measured skills, or attributes, are categorical.
Thus, these models are particularly useful when you are measuring multiple attributes that exist in different states.
For example, an educational assessment may be interested in reporting whether or not students are proficient on a set of academic standards.
Similarly, we might explore the presence or absence of attributes before and after an intervention.

There are two main classes of functions we need to get started.
Estimation functions are used for building the DCM using the [Stan probabilistic programming language](https://mc-stan.org) and getting estimates of respondent proficiency.
Evaluation functions can then be applied to the fitted model to assess how well the estimates represent the observed data.

## Installation

Because measr uses *Stan* as a backend for estimating DCMs, an installation of [rstan](https://mc-stan.org/rstan/) or [cmdstanr](https://mc-stan.org/cmdstanr/) is required.

### rstan

Before installing rstan, your system must be configured to compile C++ code.
You can find instructions on the [RStan Getting Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) guide for [Windows](https://github.com/stan-dev/rstan/wiki/Configuring-C---Toolchain-for-Windows), [Mac](https://github.com/stan-dev/rstan/wiki/Configuring-C---Toolchain-for-Mac), and [Linux](https://github.com/stan-dev/rstan/wiki/Configuring-C-Toolchain-for-Linux).

The rstan package can then be installed directly from CRAN:

```{r install-rstan, eval = FALSE}
install.packages("rstan")
```

To verify the installation was successful, you can run a test model.
If everything is set up correctly, the model should compile and sample.
For additional troubleshooting help, see the [RStan Getting Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) guide.

```{r rstan-test, eval = FALSE}
library(rstan)

example(stan_model, package = "rstan", run.dontrun = TRUE)
```

### cmdstanr

The cmdstanr package is not yet available on CRAN.
The beta release can be installed from the *Stan* R package repository:

```{r stan-repo, eval = FALSE}
install.packages("cmdstanr",
                 repos = c("https://mc-stan.org/r-packages/",
                           getOption("repos")))
```

Or the development version can be installed from [GitHub](https://github.com/stan-dev/cmdstanr):

```{r stan-dev, eval = FALSE}
# install.packages("remotes")
remotes::install_github("stan-dev/cmdstanr")
```

The cmdstanr package requires a suitable C++ toolchain.
Requirements and instructions for ensuring your toolchain is properly set up are described in the [CmdStan User Guide](https://mc-stan.org/docs/cmdstan-guide/cmdstan-installation.html#cpp-toolchain).

You can verify that the C++ toolchain is set up correctly with:

```{r check-toolchain, eval = FALSE}
library(cmdstanr)

check_cmdstan_toolchain()
```

Finally, cmdstanr requires that CmdStan (the shell interface to *Stan*).
Once the toolchain is properly set up, CmdStan can be installed with:

```{r install-cmdstan, eval = FALSE}
install_cmdstan(cores = 2)
```

For additional installation help, getting the [Getting Started with CmdStanR](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) vignette.

### measr

Once rstan and/or cmdstanr have been installed, we are ready to install measr.
The released version of measr can be installed directly from CRAN:

```{r install-measr, eval = FALSE}
install.packages("measr")
```

Or, the development version can be installed from [GitHub](https://github.com/wjakethompson/measr):

```{r measr-dev, eval = FALSE}
# install.packages("remotes")
remotes::install_github("wjakethompson/measr")
```

Once everything has been installed, we're ready to start estimating and evaluating our DCMs.

```{r load-pkg}
library(measr)
```

## Model Estimation

To illustrate, we'll fit a loglinear cognitive diagnostic model (LCDM) to an assessment of English language proficiency [see @templin-emip-2013].
There are many different subtypes of DCMs that make different assumptions about how the attributes relate to each other.
The LCDM is a general model that makes very few assumptions about the compensatory nature of the relationships between attributes.
For details on the LCDM, see @lcdm-handbook.

The data set we're using contains 29 items that together measure three attributes: morphosyntactic rules, cohesive rules, and lexical rules.
The Q-matrix defines which attributes are measured by each item.
For example, item E1 measures morphosyntactic and cohesive rules.
The data is further described in `?ecpe`.

```{r data}
ecpe_data

ecpe_qmatrix
```

We can estimate the LCDM using the `measr_dcm()` function.
We specify the data set, the Q-matrix, and the column names of the respondent and item identifiers in each (if they exist).
Finally, we add two additional arguments.
The `method` defines how the model should be estimated.
For computational efficiency, I've selected `"optim"`, which uses Stan's optimizer to estimate the model.
For a fully Bayesian estimation, you can change this `method = "mcmc"`.
Finally, we specify the type of DCM to estimate.
As previously discussed, we're estimating an LCDM in this example.
For more details and options for customizing the model specification and estimation, see the [model estimation article](https://measr.info/articles/model-estimation.html) on the measr website.

```{r est-hide, include = FALSE}
ecpe_lcdm <- measr_dcm(data = ecpe_data, qmatrix = ecpe_qmatrix,
                       resp_id = "resp_id", item_id = "item_id",
                       method = "optim", type = "lcdm",
                       file = "fits/ecpe-optim-lcdm")
```

```{r est-show, eval = FALSE}
ecpe_lcdm <- measr_dcm(data = ecpe_data, qmatrix = ecpe_qmatrix,
                       resp_id = "resp_id", item_id = "item_id",
                       method = "optim", type = "lcdm")
```

Once the model as estimated, we can use `measr_extract()` to pull out the probability that each respondent is proficient on each of the attributes.
For example, the first respondent has probabilities near 1 for all attributes, indicating a high degree of confidence that they are proficient in all attributes.
On the other hand, respondent 8 has relatively low probabilities for morphosyntactic and cohesive attributes, and is likely only proficient in lexical rules.

```{r resp-prob, message = FALSE, warning = FALSE, error = FALSE}
ecpe_lcdm <- add_respondent_estimates(ecpe_lcdm)
measr_extract(ecpe_lcdm, "attribute_prob")
```

## Model Evaluation

There are many ways to evaluate our estimated model including model fit, model comparisons, and reliability.
For a complete listing of available options, see `?model_evaluation`.
To illustrate how these functions work, we'll look at the classification accuracy and consistency metrics described by @johnson2018.

We start by adding the reliability information to our estimated model using `add_reliability()`.
We can then extract that information, again using `measr_extract()`.
For these indices, numbers close to 1 indicate a high level of classification accuracy or consistency.
These numbers are not amazing, but overall look pretty good.
For guidance on cutoff values for "good," "fair," etc. reliability, see @johnson2018.

```{r}
ecpe_lcdm <- add_reliability(ecpe_lcdm)
measr_extract(ecpe_lcdm, "classification_reliability")
```

## References
