---
title: "Estimating diagnostic classification models"
output: rmarkdown::html_vignette
bibliography: ../bib/references.bib
csl: ../bib/apa.csl
link-citations: true
vignette: >
  %\VignetteIndexEntry{Estimating diagnostic classification models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(wjake)
library(showtext)

set_theme(base_family = "Open Sans",
          plot_margin = ggplot2::margin(10, 10, 10, 10))

font_add_google("Open Sans")
showtext_auto()
showtext_opts(dpi = 192)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2916667,
  fig.align = "center",
  out.width = "90%"
)

options(mc.cores = 4,
        tidyverse.quiet = TRUE)
```

In this article, we will walk you through the steps for estimating diagnostic classification models (DCMs; also known as cognitive diagnostic models [CDMs]) using measr.
We start with the data to analyze, understand how to specify a DCM to estimate, and learn how to customize the model estimation process (e.g., prior distributions).

To use the code in this article, you will need to install and load the measr package.

```{r load-measr}
library(measr)
```

## Example Data

To demonstrate the model estimation functionality of measr, we'll examine a simulated data set.
This data set contains 2,000 respondents and 20 items that measure a total of 4 attributes, but no item measures more than 2 attributes.
The data was generated from the loglinear cognitive diagnostic model (LCDM), which is a general model that subsumes many other DCM subtypes [@lcdm].
By using a simulated data set, we can compare the parameter estimates from measr to the true data generating parameters.

```{r explore-data}
library(tidyverse)

sim_data <- read_rds("data/simulated-data.rds")

sim_data$data

sim_data$q_matrix
```

## Specifying a DCM for Estimation

In measr, DCMs are specified and estimated using the `dcm_specify()`  and `dcm_estimate()` functions, respectively.
We'll start by estimating a loglinear cognitive diagnostic model (LCDM).
The LCDM is a general DCM that subsumes many other DCM subtypes [@lcdm].

First, we specify the model using our Q-matrix (`qmatrix`).
Note that this is the only required argument to the `dcm_specify()` function.
If no other arguments are provided, sensible defaults (described below) will take care of the rest of the specification.
Next, we can specify which column, if any, in our `qmatrix` contains item identifiers.
If this variables is not present in the data, this argument can be omitted, and measr will assign identifiers based on the row number (i.e., row 1 in `qmatrix` becomes item 1).
We can then specify the type of DCM we want to estimate.
Here, we'll choose the LCDM for our measurement model, and we'll use an unconstrained structural model.^[For more details on the modeling options, see [Estimating Other DCM Sub-Types](#subtype) below.]

```{r specify-dcm}
model_spec <- dcm_specify(qmatrix = sim_data$q_matrix, identifier = "item_id",
                          measurement_model = lcdm(),
                          structural_model = unconstrained())

model_spec
```

We can then pass this model specification to `dcm_estimate()`, along with our data (`data`).
As with the model specification, we can specify which column, if any, contains respondent identifiers in the `data` object.
For the model estimation, we have the option to choose which engine to use, via the `backend` argument.
The default backend is `backend = "rstan"`, which will use the [rstan](https://mc-stan.org/rstan/) package to estimate the model.
Alternatively, we can use the [cmdstanr](https://mc-stan.org/cmdstanr/) package to estimate the model by specifying `backend = "cmdstanr"`.
The cmdstanr package works by using a local installation of *Stan* to estimate the models, rather than the version that is pre-compiled in rstan.
Once a backend has been chosen, we can supply additional arguments to those specific estimating functions.
In the example below, we specify 1,000 warm-up iterations per chain, 500 post-warm-up iterations per chain, and 4 cores to run the chains in parallel.
The full set up options available for rstan and cmdstanr can be found by looking at the help pages for `?rstan::sampling()` and `` ?cmdstanr::`model-method-sample` ``, respectively.

Finally, because estimating these models can be time intensive, we can specify a `file`.
If a file is specified, an R object of the fitted model will be automatically saved to the specified file.
If the specified `file` already exists, then the fitted model will be read back into R, eliminating the need to re-estimate the model.

```{r estimate-dcm}
lcdm <- dcm_estimate(model_spec, data = sim_data$data, identifier = "resp_id",
                     method = "mcmc", backend = "cmdstanr",
                     iter_warmup = 1000, iter_sampling = 500,
                     chains = 4, parallel_chains = 4,
                     file = "fits/sim-lcdm")
```

### Examining Parameter Estimates

Now that we've estimated a model, let's compare our parameter estimates to the true values used to generate the data.
We can start be looking at our estimates using `measr_extract()`.
This function extracts different aspects of a model estimated with measr.
Here, the `estimate` column reports estimated value for each parameter and a measure of the associated error (i.e., the standard deviation of the posterior distribution).
For example, item A1 measures two attributes and therefore has four parameters:

1. An intercept, which represents the log-odds of providing a correct response for a respondent who is proficient in neither of the attributes this item measures (i.e., `att2` and `att4`).
2. A main effect for the second attribute, which represents the increase in the log-odds of providing a correct response for a respondent who is proficient in that attribute.
3. A main effect for the fourth attribute, which represents the increase in the log-odds of providing a correct response for a respondent who is proficient in that attribute.
4. An interaction between the second and fourth attributes, which is the change in the log-odds for a respondent who is proficient in both attributes.

```{r item-params}
item_parameters <- measr_extract(lcdm, what = "item_param")
item_parameters
```

We can compare these estimates to those that were used to generate the data.
In the figure below, most parameters fall on or very close to the dashed line, which represents perfect agreement, indicating that the estimated model is accurately estimating the parameter values.

```{r lcdm-param-compare, echo = FALSE}
#| fig.asp: 0.618
#| fig.alt: >
#|   Figure shows a strong correlation between item parameters, with only a few
#|   discrepancies off of the line of perfect agreement.

library(tidyverse)
library(glue)

param_compare <- item_parameters |>
  left_join(sim_data$true_items,
            by = c("item_id", "type" = "class", "attributes",
                   "coefficient" = "coef"),
            relationship = "one-to-one") |>
  mutate(measr_est = map_dbl(estimate, mean),
         type = case_when(str_detect(coefficient, "_0") ~ "Intercept",
                          str_detect(coefficient, "_1") ~ "Main Effect",
                          str_detect(coefficient, "_2") ~ "Interaction"),
         type = factor(type, levels = c("Intercept", "Main Effect",
                                        "Interaction")))

example_item <- filter(param_compare, item_id == "A15")
example_measr <- deframe(select(example_item, attributes, measr_est))
example_true <- deframe(select(example_item, attributes, value))

msr_colors <- c("#8ECAE6", "#023047", "#D7263D")

param_compare |>
  ggplot(aes(x = measr_est, y = value)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(aes(color = type, shape = type), size = 3) +
  scale_color_manual(values = msr_colors) +
  labs(x = "measr", y = "True Value",
       color = "Parameter Type", shape = "Parameter Type")
```

We can also examine the structural parameters, which represent the overall proportion of respondents in each class.
Again, we see relatively strong agreement between the estimates from our model and the true generating values.

```{r lcdm-strc-compare, echo = FALSE}
#| fig.asp: 0.618
#| fig.alt: >
#|   Figure shows a strong correlation between item parameters, with only a few
#|   discrepancies off of the line of perfect agreement.

true_strc <- create_profiles(4) |>
  rowwise() |>
  mutate(label = glue("[{att1},{att2},{att3},{att4}]")) |>
  ungroup() |>
  select(label) |>
  mutate(value = sim_data$true_strc)

strc_compare <- measr_extract(lcdm, "strc_param") |>
  left_join(true_strc, join_by(class == label),
            relationship = "one-to-one") |>
  mutate(measr_est = map_dbl(estimate, mean))

strc_compare |>
  ggplot(aes(x = measr_est, y = value)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(size = 3, color = msr_colors[2]) +
  expand_limits(x = c(0, 0.1), y = c(0, 0.1)) +
  labs(x = "measr", y = "True Value")
```

## Customizing the Model Estimation Process

### Prior Distributions

In the code to estimate the LCDM above, we did not specify any prior distributions in the call to `measr_dcm()`.
By default, measr uses the following prior distributions for the LCDM:

```{r lcdm-prior}
default_dcm_priors(measurement_model = lcdm())
```

As you can see, main effect parameters get a `lognormal(0, 1)` prior by default.
Different prior distributions can be specified with the `prior()` function.
For example, we can specify a `normal(0, 10)` prior for the main effects with:

```{r normal-prior}
prior(normal(0, 10), type = "maineffect")
```

By default, the prior is applied to all parameters in the class (i.e., all main effects).
However, we can also apply a prior to a specific parameter.
For example, here we specify a &chi;^2^ distribution with 2 degrees of freedom as the default prior for main effects, and an exponential distribution with a rate of 2 for the main effect of attribute 1 on just item 7.
To see all parameters (including `type` and `coefficient`) that can be specified, we can pass our model specification to `get_parameters()`.

```{r multi-prior}
c(prior(chi_square(2), type = "maineffect"),
  prior(exponential(2), type = "maineffect", coefficient = "l7_11"))

get_parameters(model_spec)
```

Any distribution that is supported by the *Stan* language can be used as a prior.
A list of all distributions is available in the *Stan* documentation, and are linked to from the `?prior()` help page.

Priors can be defined before estimating the function, or created at the same time the model is estimated.
For example, both of the following are equivalent.
Here we set the prior for main effects to be a truncated normal distribution with a lower bound of 0.
This is done because the main effects in the LCDM are constrained to be positive to ensure monotonicity in the model.
Additionally note that I've set `method = "optim"`.
This means that we will estimate the model using *Stan's* optimizer, rather than using full Markov Chain Monte Carlo.
Note that the prior still influences the model when using `method = "optim"`, just as they do when using `method = "mcmc"` (the default).

```{r new-prior}
new_prior <- prior(normal(0, 15), type = "maineffect", lower_bound = 0)

new_spec <- dcm_specify(qmatrix = sim_data$q_matrix, identifier = "item_id",
                        measurement_model = lcdm(),
                        structural_model = unconstrained(),
                        priors = new_prior)

new_spec <- dcm_specify(qmatrix = sim_data$q_matrix, identifier = "item_id",
                        measurement_model = lcdm(),
                        structural_model = unconstrained(),
                        priors = prior(normal(0, 15), type = "maineffect",
                                       lower_bound = 0))

# Estimate the updated model
new_lcdm <- dcm_estimate(new_spec, data = sim_data$data, identifier = "resp_id",
                         method = "optim", backend = "cmdstanr",
                         file = "fits/sim-lcdm-optim")
```

The priors used to estimate the model are saved in the returned model object, so we can always go back and see which priors were used if we are unsure.
We can see that for the `new_lcdm` model, our specified normal prior was used for the main effects, but the default priors were still applied to the parameters for which we did not explicitly state a prior distribution.

```{r recover-prior}
measr_extract(new_lcdm, "prior")
```

### Other DCM Sub-Types {#subtype}

Although a primary motivation for measr is to provide researchers with software that makes the LCDM readily accessible, a few other popular DCM subtypes are also supported.
For example, we can estimate the deterministic inputs, noisy "and" gate [DINA, @junker2001] or the deterministic inputs, noisy "or" gate [DINO, @dino] models by specifying a different `measurement_model` in the `dcm_specify()` function.
For a complete list of currently supported measurement and structural models, see `` ?dcmstan::`measurement-model` `` and `` ?dcmstan::`structural-model` ``, respectively.

Future development work will continue to add functionality for more DCM subtypes.
If there is a specific subtype you are interested in, or would like to see supported, please open an issue on the [GitHub repository](https://github.com/wjakethompson/measr/issues).

## References
